---
title: "Weather Events Consequences For Health And Economy Across USA From 1950 To 2011"
author: "Luiz Fernando da Silva"
date: "22/05/2015"
output: html_document
---



##Synopsis

This is a simple analysis of weather events in USA and 
consequences to health of population and economy. 
The data is collected by NOAA (National Oceanic and Atmospheric 
Administration) from 1950 to november of 2011 and can be downloaded 
from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). 
For a better understanding of data format the documentation can be 
downloaded [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

With this analysis we can observe that the more harmful event 
is **TORNADO** and that the more expensive event is **FLOOD**.



##Data Processing

Assume that the dataset was downloaded and descompressed in the current working directory.

Only some columns is relevant to our analysis:

* _EVTYPE_: The type of event
* _FATALITIES_: The number of fatalities
* _INJURIES_: The number of injuries
* _PROPDMG_: Property Damage Value
* _PROPDMGEXP_: Exponent of Property Damage Value
* _CROPDMG_: Crop Damage Value
* _CROPDMGEXP_: Exponent of Crop Damage Value

We can load the data ignoring the useless columns

```{r cache=TRUE}
cols <- c('NULL','NULL','NULL','NULL','NULL','NULL','NULL',NA,'NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL',NA,NA,NA,NA,NA,NA,'NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL','NULL')
data <- read.csv('repdata-data-StormData.csv', colClasses=cols, stringsAsFactors=F)
```

Now we need to calculate the real Property Damage Value and the real Crop Damage value by yours respective exponents, looking the data we can found this following exponents:

```{r}
unique(data$PROPDMGEXP)
unique(data$CROPDMGEXP)
```

* H or h converts to hundreds (10<sup>2</sup>)
* K or k converts to thousands (10<sup>3</sup>)
* M or m converts to millions (10<sup>6</sup>)
* B or b converts to billions (10<sup>9</sup>)
* Blank values keeps single values
* Other numeric converts to respective exponent (10<sup>x</sup>)
* Other symbols will be ignored

```{r}
data$PROPDMGEXPN[data$PROPDMGEXP == 'H'] <- 1e+02
data$PROPDMGEXPN[data$PROPDMGEXP == 'h'] <- 1e+02
data$PROPDMGEXPN[data$PROPDMGEXP == 'K'] <- 1e+03
data$PROPDMGEXPN[data$PROPDMGEXP == 'M'] <- 1e+06
data$PROPDMGEXPN[data$PROPDMGEXP == 'm'] <- 1e+06
data$PROPDMGEXPN[data$PROPDMGEXP == 'B'] <- 1e+09
data$PROPDMGEXPN[data$PROPDMGEXP == ''] <- 1e+00
data$PROPDMGEXPN[data$PROPDMGEXP == '0'] <- 1e+00
data$PROPDMGEXPN[data$PROPDMGEXP == '1'] <- 1e+01
data$PROPDMGEXPN[data$PROPDMGEXP == '2'] <- 1e+02
data$PROPDMGEXPN[data$PROPDMGEXP == '3'] <- 1e+03
data$PROPDMGEXPN[data$PROPDMGEXP == '4'] <- 1e+04
data$PROPDMGEXPN[data$PROPDMGEXP == '5'] <- 1e+05
data$PROPDMGEXPN[data$PROPDMGEXP == '6'] <- 1e+06
data$PROPDMGEXPN[data$PROPDMGEXP == '7'] <- 1e+07
data$PROPDMGEXPN[data$PROPDMGEXP == '8'] <- 1e+08
data$PROPDMGEXPN[data$PROPDMGEXP == '+'] <- 0
data$PROPDMGEXPN[data$PROPDMGEXP == '?'] <- 0
data$PROPDMGEXPN[data$PROPDMGEXP == '-'] <- 0
data$PROPDMGV <- data$PROPDMG * data$PROPDMGEXPN

data$CROPDMGEXPN[data$CROPDMGEXP == 'K'] <- 1e+03
data$CROPDMGEXPN[data$CROPDMGEXP == 'k'] <- 1e+03
data$CROPDMGEXPN[data$CROPDMGEXP == 'M'] <- 1e+06
data$CROPDMGEXPN[data$CROPDMGEXP == 'm'] <- 1e+06
data$CROPDMGEXPN[data$CROPDMGEXP == 'B'] <- 1e+09
data$CROPDMGEXPN[data$CROPDMGEXP == ''] <- 1e+00
data$CROPDMGEXPN[data$CROPDMGEXP == '0'] <- 1e+00
data$CROPDMGEXPN[data$CROPDMGEXP == '2'] <- 1e+02
data$CROPDMGEXPN[data$CROPDMGEXP == '?'] <- 0
data$CROPDMGV <- data$CROPDMG * data$CROPDMGEXPN
```

Now we can start analysing the data.


##Results

####Question 1. Across the United States, which types of events are most harmful with respect to population health?
Look at top 10 most harmful weather events for public health.

```{r cache=TRUE}
total_fatalities <- aggregate(FATALITIES ~ EVTYPE, data=data, sum)
total_injuries <- aggregate(INJURIES ~ EVTYPE, data=data, sum)
total1 <- merge(total_fatalities, total_injuries, by='EVTYPE')
ranking1 = total1[order(-total1$FATALITIES, -total1$INJURIES),][1:10,]

par(mfrow=c(2,1), cex = 0.6)
with(ranking1,barplot(FATALITIES, names.arg=EVTYPE, col='Red', las=2, main="Top 10 Most Harmful Events", ylab="Fatalities"))
with(ranking1,barplot(INJURIES, names.arg=EVTYPE, col='Orange', las=2, ylab="Injuries"))
```

```{r}
# Show first 5 items
ranking1[1:5,]
```


####Question 2. Across the United States, which types of events have the greatest economic consequences?
Look at top 10 most expensive weather events in billions of dollars.

```{r cache=TRUE}
total_prop <- aggregate(PROPDMGV ~ EVTYPE, data=data, sum)
total_prop$PROPDMGV <- total_prop$PROPDMGV / (10^9)
total_crop <- aggregate(CROPDMGV ~ EVTYPE, data=data, sum)
total_crop$CROPDMGV <- total_crop$CROPDMGV / (10^9)
total2 <- merge(total_prop, total_crop, by='EVTYPE')
ranking2 = total2[order(-total2$PROPDMGV, -total2$CROPDMGV),][1:10,]

par(mfrow=c(2,1), cex = 0.6)
with(ranking2,barplot(PROPDMGV, names.arg=EVTYPE, col='Gold', las=2, main="Top 10 Most Expensive Events ($ Billions)", ylab="Property Damage"))
with(ranking2,barplot(CROPDMGV, names.arg=EVTYPE, col='LightGrey', las=2, ylab="Crop Damage"))
```


```{r}
# Show first 5 items
ranking2[1:5,]
```